\ZpL: a p-adic precision package

### 2018-02-04 ZpL/ZpL.tex
Adding summary of paper

\author{Xavier Caruso}
  \affiliation{Universit\'e Rennes 1; \\
  %\institution{IRMAR}
  %\city{Rennes, France}
  %\postcode{35042}
}
\email{xavier.caruso@normalesup.org}
\author{David Roe}
  \affiliation{MIT; \\
  %\institution{Department of Mathematics}
  %\city{Pittsburgh, PA, USA}
  %\postcode{15260}
}
\email{roed@mit.edu}
\author{Tristan Vaccon}
  \affiliation{Universit\'e de Limoges; \\
  %\institution{CNRS, XLIM UMR 7252}
  %\city{Limoges, France}  
  %\postcode{87060}  
  }
\email{tristan.vaccon@unilim.fr}

@@

\begin{abstract}
We present a new package \ZpL for the Computer Algebra system \sage. It 
implements a sharp tracking of precision on $p$-adic numbers, following 
the theory of ultrametric precision introduced in \cite{caruso-roe-vaccon:14a}. The 
underlying algorithms are mostly based on Automatic Differentiation 
techniques. We introduce them, study their complexity and discuss our 
design choices.
We illustrate the benefits of our package (in comparison with previous 
implementations) with a large sample of examples coming from Linear 
Algebra, Commutative Algebra and Differential Equations.
\end{abstract}

@@

The standard method for handling precision when computing with real numbers is floating point arithmetic, which may also be used in $p$-adic computation.  At a given precision level, a finite set of representable numbers are chosen, and arithmetic operations are defined to give a representable number that is close to the true result \cite{ieee754:2008}.  Floating point arithmetic has the benefit of efficient arithmetic operations, but users are responsible for tracking the precision of the results.  Numerically unstable algorithms can lead to very inaccurate answers \todo{add reference for numerical stability}.

If provably correct results are desired, interval arithmetic provides an alternative to floating point.  Instead of just tracking an approximation to the answer, the package also tracks a radius within which the true result lies.  This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.  Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision, (see Section \ref{sec:demo} for many examples).

For better control of precision, we may use ellipsoids rather than balls.  This change would have little utility for real numbers, since ellipsoids are not preserved by most functions.  For $p$-adic fields, in contrast, sufficiently small ellipsoids are preserved by differentiable maps.  From an algebraic perspective, these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space, and Proposition \ref{prop:precision} describes how the image of such a coset under a map $f$ is given exactly by applying the differential of $f$ to $H$.

In this paper, we describe an implementation of this idea in \sage \cite{sage}.  \todo{Add short description of package features.}

\todo{Plan of article}

\todo{Convention: in matrices, vectors are represented by row-vector.}

@@

\noindent
\ZpXX is a generic notation for \ZpCR, \ZpLC and \ZpLF.
The first one \ZpCR is the usual constructor for $p$-adic parents in 
\sage. It tracks precision using interval arithmetics. On the contrary
\ZpLC and \ZpLF are provided by our package. In the sequel, we will
compare the results output with the three 
Results for \ZpLF are only shown when they differ from \ZpLC.

\subsection{Elementary arithmetic}

@@

\noindent
Multiplication by $p$ (here $3$) is a shift on the digits and thus
leads to a gain of one digit in absolute precision.
On the example below, we observe that when this multiplication is
split into several steps, \ZpCR does not see the gain of precision
while \ZpLC and \ZpLF do.

@@

\subsection{Linear algebra}

@@

\subsection{Commutative algebra}

@@

\subsection{$p$-adic differential equations}

@@

\section{What's behind the scene?}

@@

This Section is organized as follows.
\S \ref{ssec:preclemma} contains a brief overview on the theory of
$p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
In the next two subsections, we explain in more details how \ZpLC and 
\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
Automatic Differentiation leading to the actual computation of a
global object modeling the precision (which is the aforementioned 
lattice). 
In \S \ref{ssec:viewprec}, we explain how precision on any individual number
can be recoved from the above computations and discuss the validity
of our results.
The overhead complexity induced by our package is analyzed in
\S \ref{ssec:complexity}.

@@

In the case of \ZpLF, creating a variable has a cost $O(s_\inp)$ 
whereas deleting a variable is free. 
Thus when executed with the \ZpLF mecanism, the complexity of our
algorithm becomes $O(s_\inp c)$.

@@

\section{Conclusion}

\begin{itemize}
\item In which case should we use this package? Mainly for exploration
\item example of $p$-adic differential equations (especially for $p=2$)
\end{itemize}

\subsection{Exploration}

>>>>

\author{Xavier Caruso}
  \affiliation{Universit\'e Rennes 1; \\
  %\institution{IRMAR}
  %\city{Rennes, France}
  %\postcode{35042}
}
\email{xavier.caruso@normalesup.org}
\author{David Roe}
  \affiliation{MIT; \\
  %\institution{Department of Mathematics}
  %\city{Cambridge, MA, USA}
  %\postcode{02139}
}
\email{roed@mit.edu}
\author{Tristan Vaccon}
  \affiliation{Universit\'e de Limoges; \\
  %\institution{CNRS, XLIM UMR 7252}
  %\city{Limoges, France}  
  %\postcode{87060}  
  }
\email{tristan.vaccon@unilim.fr}

@@

\begin{abstract}
We present a new package \ZpL for the mathematical software system \sage. It 
implements a sharp tracking of precision on $p$-adic numbers, following 
the theory of ultrametric precision introduced in \cite{caruso-roe-vaccon:14a}. The 
underlying algorithms are mostly based on automatic differentiation 
techniques. We introduce them, study their complexity and discuss our 
design choices.
We illustrate the benefits of our package (in comparison with previous 
implementations) with a large sample of examples coming from linear 
algebra, commutative algebra and differential equations.
\end{abstract}

@@

The standard method for handling precision when computing with real numbers
is floating point arithmetic, which may also be used in $p$-adic computation.
At a given precision level, a finite set of representable numbers are chosen,
and arithmetic operations are defined to give a representable number
that is close to the true result \cite{ieee754:2008}.  Floating point arithmetic
has the benefit of efficient arithmetic operations, but users are responsible
for tracking the precision of the results.  Numerically unstable algorithms can
lead to very inaccurate answers \cite{higham:2002}.

If provably correct results are desired, interval arithmetic provides
an alternative to floating point.  Instead of just tracking an approximation to the answer,
the package also tracks a radius within which the true result lies.
This method is commonly used for $p$-adic computations since
the ultrametric property of $p$-adic fields frequently keeps the radius small.
Computations remain fairly efficient with this approach, but numerical instability
can still lead to dramatic losses in precision, (see Section \ref{sec:demo} for many examples).
Tracking the precision of multiple variables concurrently, the set of possible
true values associated to an inexact value takes the form of an ellipsoid
with axes parallel to the coordinate axes.

For better control of precision, we may allow arbitrary axes.  This change would have
little utility for real numbers, since such ellipsoids are not preserved by most functions.
For $p$-adic fields, in contrast, sufficiently small ellipsoids are mapped to other
ellipsoids by differentiable maps.  From an algebraic perspective,
these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space,
and Proposition \ref{prop:precision} describes how the image of such a coset
under a map $f$ is given exactly by applying the differential of $f$ to $H$.

In this paper, we describe an implementation of this idea in \sage \cite{sage}.
Rather than attaching a precision to each element, we store the precision of
many elements together by tracking a precision lattice for the whole collection
of variables.  As variables are created and destroyed, we update a matrix whose
rows represent the vectors in the lattice.  Information about the precision
of elements is extracted from the matrix as necessary.

The article is structured as follows. In \S \ref{sec:demo} we provide a demonstration
of the package, showing how it can provide more precise answers than the
traditional methods for tracking $p$-adic precision.  In particular, \S \ref{ssec:elem}
describes elementary arithmetic and the SOMOS-4 sequence, \S \ref{ssec:linalg}
gives examples from linear algebra, \S \ref{ssec:comalg} examples using polynomials,
and \S \ref{ssec:diffeq} examples of differential equations.

In \S \ref{sec:implementation} we give more details on the implementation.
\S \ref{ssec:preclemma} contains a brief overview on the theory of
$p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
In the next two subsections, we explain in more details how \ZpLC and 
\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
automatic differentiation leading to the actual computation of the
lattice that models the precision.
In \S \ref{ssec:viewprec}, we explain how precision on any individual number
can be recovered and discuss the validity
of our results.
The complexity overhead induced by our package is analyzed in
\S \ref{ssec:complexity}.

Finally, \S \ref{sec:conclusion} contains a discussion of how we see this
package fitting into the existing $p$-adic implementations.  While these
methods do introduce overhead, they are well suited to exploring
precision behavior when designing algorithms, and can provide
hints as to when further precision analysis would be useful.

@@

\noindent
\ZpXX is a generic notation for \ZpCR, \ZpLC and \ZpLF.
The first one \ZpCR is the usual constructor for $p$-adic parents in 
\sage. It tracks precision using interval arithmetic. On the contrary
\ZpLC and \ZpLF are provided by our package. In the sequel, we will
compare the results output with the three 
Results for \ZpLF are only shown when they differ from \ZpLC.

\subsection{Elementary arithmetic}
\label{ssec:elem}

@@

\noindent
Multiplication by $p$ (here $3$) is a shift on the digits and thus
leads to a gain of one digit in absolute precision.
In the example below, we observe that when this multiplication is
split into several steps, \ZpCR does not see the gain of precision
while \ZpLC and \ZpLF do.

@@

\subsection{Linear algebra}
\label{ssec:linalg}

@@

\subsection{Commutative algebra}
\label{ssec:comalg}

@@

\subsection{$p$-adic differential equations}
\label{ssec:diffeq}

@@

\section{What's behind the scene?}
\label{sec:implementation}

@@

@@

In the case of \ZpLF, creating a variable has a cost $O(s_\inp)$ 
whereas deleting a variable is free. 
Thus when executed with the \ZpLF mechanism, the complexity of our
algorithm becomes $O(s_\inp c)$.

@@

\section{Conclusion}
\label{sec:conclusion}

%\begin{itemize}
%\item In which case should we use this package? Mainly for exploration
%\item example of $p$-adic differential equations (especially for $p=2$)
%\end{itemize}

%\subsection{Exploration}
